Fast gravitational wave parameter estimation without compromises

### 2023-02-07 src/tex/ms.tex
new plotting and table rules

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{color}
%\usepackage[dvipsnames]{xcolor}

@@

% Brief description of PE
Parameter estimation (PE) underpins all of gravitational-wave physics and
astrophysics, and is one of the most commonly performed tasks in
gravitational-wave (GW) data analysis \citep{Christensen:2022bxb,
2019PASA...36...10T}. The central goal of PE is to infer the parameters of a
particular GW source given the strain data recorded by instruments like LIGO
\citep{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp}, Virgo \citep{VIRGO:2014yos} and KAGRA
\citep{KAGRA:2020tym}.  In the standard compact binary coalescence (CBC)
scenario, this could mean inferring intrinsic parameters such as the masses and
spins of the compact objects, as well as extrinsic parameters such as their sky
localization and distance from Earth. PE is also applied to test general
relativity (GR)
\citep{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio},
and constrain the properties of nuclear matter \citep{LIGOScientific:2018hze}.
PE is a crucial step in GW science, since it translates characteristics of the
strain data into astrophysically relevant quantities that can be used to
constrain astrophysical phenomena, including informing theories of binary
evolution \citep{LIGOScientific:2021psn}.

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\textsc{LALInference} \citep{Veitch:2014wba}, \textsc{PyCBC Inference}
\citep{Biwer:2018osg}, and \textsc{Bilby}
\citep{Ashton:2018jfp,Romero-Shaw:2020owr}.  These packages have been tested by a
number of groups and are well regarded as standard tools. However, while these
tools have passed many robustness tests, they are known to be computationally
intensive. The exact amount of time needed to analyze one event depends on
factors like the duration and frequency of the signal, as well as features of
the specific waveform model. Typical runtimes for production-level analyses can
range from hours to weeks. This expense precludes iterating quickly on results,
launching large scale measurement simulations, or obtaining results in low
latency to inform astronomers for potential followup in real time.

Additionally, in the coming decade, there are planned upgrades for existing
facilities, as well as plans for next-generation detectors such as the Einstein
Telescope (ET) \citep{Punturo:2010zz} and the Cosmic Explorer (CE)
\citep{LIGOScientific:2016wof}. These upgrades will increase sensitivity of
the instruments and allow for the detection of more events with a better
signal-to-noise ratio (SNR). The number of events that will be detected in the
coming decade is expected to grow from around a thousand per year to over a
million per year \citep{Baibhav:2019gxm}. This will put a significant strain on
the current PE tools.

% Effort going to new tools development.
In order to address this, there are efforts from multiple groups to speed up the
PE process. This includes methods that employ techniques such as
reduced-order quadrature \citep{Canizares:2014fya,Smith:2016qas}, adaptive proposal distributions in nested sampling
\citep{Williams:2021qyt}, deep learning networks pre-trained on large collections
of waveforms \citep{Dax:2021tsq,Dax:2022pxd}, as well as methods that reduce the
computational expense of classical PE by leveraging our knowledge of GW signals
\citep{Veitch:2014wba,Ashton:2021anp,Cornish:2021wxy,Islam:2022afg,Roulet:2022kot,Lee:2022jpn,Lange:2018pyp,Wofford:2022ykb}.
While these avenues are promising for standard GW problems, they rely on
assumptions that may not hold for analyses targeting additional physical effects
beyond standard CBCs in GR, such as matter effects, lensing and deviations from GR.

@@

To compute the integral in Eq.~\eqref{eq: innerproduct}, we need to evaluate a
waveform model $h(\mathbf{\theta})$ at a number of frequency bins. This makes
evaluating the likelihood function often the most computationally intensive
part of PE. The most accurate waveforms are obtained via numerical relativity
(NR) simulations \citep{BaumgarteShaprio:NumRel}, which directly solve the Einstein equations numerically for a
given system. Unfortunately, such simulations can take anywhere from a day to
half a year, making direct evaluation through NR prohibitively expensive. To
circumvent this, there are several families of waveform ``approximants'',
including the \textsc{IMRPhenom} family \citep{Khan:2015jqa,
Garcia-Quiros:2020qpx}, the \textsc{SEOB} family \citep{PhysRevD.89.061502}, and
the NR surrogate family \citep{Varma:2019csw}. Evaluating those approximants can
take ${\sim}10$ ms to 1 s for short signals, such as a $30-30\ M_{\odot}$
binary black hole, reaching up to \kw{Fill} for long signals, such as a
$1.4-1.4\ M_{\odot}$ binary neutron star.  Since PE requires millions of
likelihood evaluations during sampling, the computational cost in evaluating
the waveform is a major contributor to the long runtimes of GW PE.

@@

Since the cost of evaluating a waveform model scales linearly with the number
of time or frequency bins, the computational burden for longer-duration signals
is often quite large. To mitigate this, there are a number of methods to reduce
the number of basis points needed to compute the likelihood faithfully
\citep{Field:2011mf, Field:2013cfa, Smith:2016qas,
Vinciguerra:2017ngf,Morisaki:2020oqk,Morisaki:2021ngj}. In this work, we take
advantage of likelihood heterodyning \citep{Cornish:2010kf,Cornish:2021lje}
(also known as relative binning \citep{Zackay:2018qdy}).

@@

To obtain a reference waveform, we currently use the \textsc{differential
evolution} algorithm \citep{Storn1997DifferentialE} available in the
\textsc{scipy} package \citep{2020SciPy-NMeth} to find the waveform parameters
which maximize the likelihood.  The reference waveform could also be produced
from trigger parameters precomputed by a search pipeline without additional
computation. Once we have obtained a reference waveform, we can check the accuracy
of the heterodyned likelihood by comparing its value to the original likelihood
at several points in the parameter space. We can choose the number of bins such
that the difference between the values of the two likelihoods is smaller than
chosen tolerance threshold.

@@

Given Eq.~\eqref{eq: loglikelihood} and the prior, one can evaluate the
posterior density, Eq.~\eqref{eq: bayes}, over the entire parameter space of
interest to obtain the most probable set of values that are consistent with the
data. However, directly sampling this posterior quickly becomes intractable as
the dimensionality of the parameter space increases beyond a few dimensions.
Markov chain Monte Carlo \citep{gelmanbda04} is a common method employed to
generate samples from the target posterior when direct sampling is not
possible.

In MCMC, the posterior distribution is approximated by a Markov chain that
eventually converges to the target distribution \citep{10.1214/aos/1176325750}.
The chain is constructed by iteratively proposing a new point in the
parameter space based on the current location of the chain. The proposed point
is accepted with a probability that is usually set to be proportional to the
ratio of the posterior density evaluated at the proposed point and the current
point. The chain can either accept the proposal and move to the new location,
or reject the proposal and stay at the current location. This process is
repeated until the chain converges to the target distribution. The samples
generated by the chain are then used as a fair sample to estimate the
quantities of interest, such as the mean and credible intervals of the source
parameters. In practice, since we do not know the target distribution ahead of
time, the MCMC process is usually repeated until a certain criterion is met,
such as a Gelman-Rubin convergence statistic \citep{10.2307/2246093} lower than
certain threshold, or simply after a fixed number of iterations.

@@

However, as we often deal with high-dimensional problems, even the optimally
tuned Gaussian transition kernel does not guarantee good performance. In order
to have a reasonable acceptance rate, the variance of the Gaussian has to be
smaller in a higher dimensional space, which means that the transition kernel
will generally make smaller and smaller steps as we increase the dimensionality
of the problem \citep{2017arXiv170102434B}.

Transition kernels that leverage gradient information of the target
distribution can help address this issue of shortening steps in a high
dimensional space. Instead of proposing a new point by drawing from a Gaussian,
one can use the gradient evaluated at the current location to propose a new
point, so that the evolution of the chain is preferentially directed to regions
of higher probability. For example, the Metropolis-adjusted Langevin algorithm
(MALA) \citep{10.2307/2346184} places a unit Gaussian at the tip of the gradient
vector at the current position,
\begin{align}
    \mathbf{x} = \mathbf{x_0} + \tau \nabla\log{p(\mathbf{x_0})} + \sqrt{2\tau}N(0,\mathbf{I})\, ,
\end{align}
where $\tau$ is a step size chosen during the tuning stage. Compared to a
Gaussian centered at the current location, the MALA transition kernel is more
likely to propose a point in the higher posterior density region because of the
gradient term, which helps boost the acceptance rate.
Hamiltonian Monte Carlo \citep{2017arXiv170102434B} is another gradient-based
algorithm that has been explored for neutron star inspirals
\citep{Bouffanais:2018hoz}.

While transition kernels that use gradient information can help improve the
acceptance rate, computing the gradient of the posterior density function
introduces an additional computational cost, which is not necessarily
beneficial in terms of sampling time. If one wants to compute the gradient
through finite differencing, the additional computational cost goes as at least
${\sim} \mathcal{O}(2n)$, where $n$ is the dimension of the problem. On the
other hand, schemes like \textsc{Jax}
\citep{jax2018github} allow us to compute the gradient of the likelihood
function with respect to the parameters through automatic differentiation,
which gives the gradient information down to machine precision at around the
same order of time compared to evaluating the posterior itself. Thus, having
access to gradient information through automatic differentiation is crucial to
making gradient-based transition kernels favorable in terms of computing cost.

To leverage automatic differentiation, the entire posterior function must be
implemented within \textsc{Jax} or a similar framework; this includes the
likelihood and, therefore, the waveform approximant.  This means that the
development of differentiable approximants is essential for leveraging
gradient-based sampling in GW applications. We currently make use of the
waveforms implemented in \textsc{ripple} \citep{ripplepaper}.

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to outperform gradient-free
algorithms in many practical applications, there remain classes of problems
that most gradient-based samplers do not solve well. For example, first-order
gradient-based algorithms struggle with target distributions that exhibit
locally-varying correlations, since they assume a single mass matrix that does
not depend on the location of the chain by construction
\citep{2017arXiv170102434B}.\footnote{Sampling algorithms that use higher order
derivatives such as manifold-MALA and Riemannian-HMC \citep{RMHMC} can in
principle handle local correlations in the target distribution; however, they
often encounter instabilities when used in real-life applications, so their use
is a rare practice.} Another example is multi-modality: if there are multiple
modes in the target distribution, individual chains will likely be trapped in
one mode and take an extremely long time to transverse between the modes
\citep{2018arXiv180803230M}.  This means that the relative weights between modes
will take much longer to estimate than the shape of each mode.

@@

% Crux of normalizing flow
All the above issues can be mitigated by normalizing flows.  Normalizing flows
are a technique based on neural networks that aims at learning a mapping from a
simple distribution, such as a Gaussian, to a complex distribution, often given
in the form of samples \citep{2019arXiv190809257K, 2019arXiv191202762P}. Once
the network is trained, one can evaluate the probability density of the complex
distribution and sample from it very efficiently, by first evaluating the
simple distribution and then applying the learned mapping. The core equation
of normalizing flows is the coordinate transformation of probability
distributions via a Jacobian, as given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|^{-1},
\end{align}
where $p_x(X)$ is the complex target distribution, $p_z(Z)$ is the simple
latent distribution and $f$ is an invertible parameterized transform that
connects the two distributions, $x = f(z)$, to be learned by the normalizing
flow. See \cite{2019arXiv190809257K, 2019arXiv191202762P} for a detailed
discussion of the algorithm.

@@

Note that every time we retrain the network, we are breaking the Markov
properties since we are changing the proposal distribution. To produce final
samples that can be used to estimate target quantities, one has to freeze the
normalizing flow model and not retrain during the final sampling phase in order
to satisfy the detailed balance condition. We use the package \textsc{flowMC}
\citep{2022arXiv221106397W,Gabrie:2021tlu}, with MALA as the gradient-based sampler. The
pseudocode of the algorithm is given in Algorithm \ref{alg:cap}.

@@

To demonstrate the robustness of our pipeline, we use it to recover the
parameters of a set of simulated signals injected into different instantiations
of synthetic stationary Gaussian noise. Then we run our pipeline on the
simulated data, and determine the credible interval at which the true
parameters of the injected signals are recovered. From the set of credible
values, we can check whether the truth lies within a certain credible interval
at the expected frequency: if our pipeline is working as expected, we should
find that the true parameters lie within $x\%$ credible interval $x\%$ of the
time, e.g., the true value should lie within the $50\%$ credible interval
$50\%$ of the time. In other words, the recovered percentiles of the true
parameters should be uniformly distributed. Deviation from this behavior would
suggest the pipeline is either over-confident or too conservative
\citep{Cook2006,Talts2018}.

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over 16 s of data, with a minimum frequency
cutoff of 30 Hz and a sampling rate of 2048 Hz. We draw noise from a set of
projected design PSDs for the LIGO Hanford, LIGO Livingston
(\texttt{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo
(\texttt{SimNoisePSDAdvVirgo}) detectors
\citep{lalsuite,Shoemaker:T0900288,2012arXiv1202.4031M}. For both injection and
recovery, we make use of the \textsc{IMRPhenomD} waveform \citep{Khan:2015jqa}
via the fully-differentiable implementation presented in the \textsc{ripple}
package \citep{ripplepaper}. We use a neural spline flow model \citep{2019arXiv190604032D}
with 10 layers, each with 128 hidden units, and 8 bins per layer as our
normalizing flow model.

@@

To further quantify how well our result agrees with a uniform distribution, we
can compute the Kolmogorov-Smirnov $p$-values for each marginalized
distribution \citep{}.  A low $p$-value (with a threshold often
chosen to be $p = 0.05$) could indicate that our result is in tension with a uniform
distribution. The $p$-values obtained for each parameter are shown in the
legend of Fig.~\ref{fig:ppplot}. Most of them are well above the $p = 0.05$
threshold, except for $\psi$, which is close to the threshold. Once
again, assuming these $p$-values are drawn from a uniform distribution, given
11 draws (the number of parameters in our inference), it is not abnormal to
have one of the parameters lying slightly outside the threshold. To assess
whether this is expected, we can compute the combined $p$-value for these 11
parameters, and find it to be $p = 0.70$.  This shows our inference pipeline
performs properly on simulated data at a similar level as standard tools
\citep{Veitch:2014wba,Romero-Shaw:2020owr}.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
starting at 20 Hz for GW150914, and 128 s of data sampled at 4096 Hz starting
at 23 Hz for GW170817; strain data and PSDs for both events are fetched from
GWOSC \citep{GWOSC}. We use the same normalizing flow model as the
injection-recovery study. For our specific choice of sampler settings, we
produce ${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using \textsc{arviz}
\citep{arviz_2019}, \url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
} for GW150914 and GW170817 respectively. Running on an \textsc{Nvidia} A100
GPU, the wall time for both events is around 3 minutes. Most of this time is
spent on just-in-time (JIT) compilation of the code; the actual sampling time
is only ${\sim}40$ s. We pre-compute the reference waveform parameters used to
hetetrodyne the likelihood for the two events, which is omitted in the
wall-time calculation.

\begin{figure}
    \script{plotEvents.py}
    \includegraphics[width=0.99\linewidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (blue) and \textsc{Bilby} (orange).
    }
    \label{fig:GW150914}
\end{figure}

\begin{figure}
    \script{plotEvents.py}
\includegraphics[width=0.99\linewidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (blue) and \textsc{Bilby} (orange).
}
\label{fig:GW170817}
\end{figure}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files included in our data release \citep{release}.  We
carry out these runs using \textsc{parallel Bilby} (\textsc{pBilby})
\citep{Smith:2019ucc} to distribute the computation over 400 Intel Skylake CPUs
for each event.  For the specific settings chosen, the wall-time duration of
each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day for GW170817.

@@

\begin{table}[hbt!]
    \begin{center}
    \begin{tabular}{ l l l}
    \hline
    \hline
    Parameter & GW150914 & GW170817\\
    \hline
        $M_c$ & $\textbf{0.0171}$ & $\textbf{0.0166}$ \\
        $q$ & $0.0036$ & $0.0004$ \\
        $\chi_1$ & $0.0023$ & $0.0006$ \\
        $\chi_2$ & $0.0098$ & $0.0004$ \\
        $d_L$ & $0.0009$ & $0.0076$ \\
        $\phi_c$ & $0.0003$ & $0.0024$ \\
        $\iota$ & $0.0026$ & $0.0045$ \\
        $\psi$ & $0.0005$ & $0.0034$ \\
        $\alpha$ & $0.0007$ & $0.0242$ \\
        $\delta$ & $0.0020$ & $0.0017$ \\
    \hline
    \hline
    \end{tabular}
    \caption{JSD values for the marginalized distribution of each parameter for
    GW150914 and GW170817 between our code and \textsc{Bilby}. The bold values
    indicate the parameters with the largest JSD.}
    \label{tab:JSD}
    \end{center}
\end{table}

>>>>

% Define document class
\documentclass[twocolumn]{aastex631}
\usepackage{showyourwork}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usepackage{algorithmic}
\usepackage{color}
\usepackage{booktabs}

@@

% Brief description of PE
Parameter estimation (PE) underpins all of gravitational-wave physics and
astrophysics, and is one of the most commonly performed tasks in
gravitational-wave (GW) data analysis \cite{Christensen:2022bxb,
2019PASA...36...10T}. The central goal of PE is to infer the parameters of a
particular GW source given the strain data recorded by instruments like LIGO
\cite{LIGOScientific:2014pky,LIGOScientific:2021usb,LIGOScientific:2021djp}, Virgo \cite{VIRGO:2014yos} and KAGRA
\cite{KAGRA:2020tym}.  In the standard compact binary coalescence (CBC)
scenario, this could mean inferring intrinsic parameters such as the masses and
spins of the compact objects, as well as extrinsic parameters such as their sky
localization and distance from Earth. PE is also applied to test general
relativity (GR)
\cite{LIGOScientific:2016lio,LIGOScientific:2018dkp,LIGOScientific:2021sio},
and constrain the properties of nuclear matter \cite{LIGOScientific:2018hze}.
PE is a crucial step in GW science, since it translates characteristics of the
strain data into astrophysically relevant quantities that can be used to
constrain astrophysical phenomena, including informing theories of binary
evolution \cite{LIGOScientific:2021psn}.

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\textsc{LALInference} \cite{Veitch:2014wba}, \textsc{PyCBC Inference}
\cite{Biwer:2018osg}, and \textsc{Bilby}
\cite{Ashton:2018jfp,Romero-Shaw:2020owr}.  These packages have been tested by a
number of groups and are well regarded as standard tools. However, while these
tools have passed many robustness tests, they are known to be computationally
intensive. The exact amount of time needed to analyze one event depends on
factors like the duration and frequency of the signal, as well as features of
the specific waveform model. Typical runtimes for production-level analyses can
range from hours to weeks. This expense precludes iterating quickly on results,
launching large scale measurement simulations, or obtaining results in low
latency to inform astronomers for potential followup in real time.

Additionally, in the coming decade, there are planned upgrades for existing
facilities, as well as plans for next-generation detectors such as the Einstein
Telescope (ET) \cite{Punturo:2010zz} and the Cosmic Explorer (CE)
\cite{LIGOScientific:2016wof}. These upgrades will increase sensitivity of
the instruments and allow for the detection of more events with a better
signal-to-noise ratio (SNR). The number of events that will be detected in the
coming decade is expected to grow from around a thousand per year to over a
million per year \cite{Baibhav:2019gxm}. This will put a significant strain on
the current PE tools.

% Effort going to new tools development.
In order to address this, there are efforts from multiple groups to speed up the
PE process. This includes methods that employ techniques such as
reduced-order quadrature \cite{Canizares:2014fya,Smith:2016qas}, adaptive proposal distributions in nested sampling
\cite{Williams:2021qyt}, deep learning networks pre-trained on large collections
of waveforms \cite{Dax:2021tsq,Dax:2022pxd}, as well as methods that reduce the
computational expense of classical PE by leveraging our knowledge of GW signals
\cite{Veitch:2014wba,Ashton:2021anp,Cornish:2021wxy,Islam:2022afg,Roulet:2022kot,Lee:2022jpn,Lange:2018pyp,Wofford:2022ykb}.
While these avenues are promising for standard GW problems, they rely on
assumptions that may not hold for analyses targeting additional physical effects
beyond standard CBCs in GR, such as matter effects, lensing and deviations from GR.

@@

To compute the integral in Eq.~\eqref{eq: innerproduct}, we need to evaluate a
waveform model $h(\mathbf{\theta})$ at a number of frequency bins. This makes
evaluating the likelihood function often the most computationally intensive
part of PE. The most accurate waveforms are obtained via numerical relativity
(NR) simulations \cite{BaumgarteShaprio:NumRel}, which directly solve the Einstein equations numerically for a
given system. Unfortunately, such simulations can take anywhere from a day to
half a year, making direct evaluation through NR prohibitively expensive. To
circumvent this, there are several families of waveform ``approximants'',
including the \textsc{IMRPhenom} family \cite{Khan:2015jqa,
Garcia-Quiros:2020qpx}, the \textsc{SEOB} family \cite{PhysRevD.89.061502}, and
the NR surrogate family \cite{Varma:2019csw}. Evaluating those approximants can
take ${\sim}10$ ms to 1 s for short signals, such as a $30-30\ M_{\odot}$
binary black hole, reaching up to \kw{Fill} for long signals, such as a
$1.4-1.4\ M_{\odot}$ binary neutron star.  Since PE requires millions of
likelihood evaluations during sampling, the computational cost in evaluating
the waveform is a major contributor to the long runtimes of GW PE.

@@

Since the cost of evaluating a waveform model scales linearly with the number
of time or frequency bins, the computational burden for longer-duration signals
is often quite large. To mitigate this, there are a number of methods to reduce
the number of basis points needed to compute the likelihood faithfully
\cite{Field:2011mf, Field:2013cfa, Smith:2016qas,
Vinciguerra:2017ngf,Morisaki:2020oqk,Morisaki:2021ngj}. In this work, we take
advantage of likelihood heterodyning \cite{Cornish:2010kf,Cornish:2021lje}
(also known as relative binning \cite{Zackay:2018qdy}).

@@

To obtain a reference waveform, we currently use the \textsc{differential
evolution} algorithm \cite{Storn1997DifferentialE} available in the
\textsc{scipy} package \cite{2020SciPy-NMeth} to find the waveform parameters
which maximize the likelihood.  The reference waveform could also be produced
from trigger parameters precomputed by a search pipeline without additional
computation. Once we have obtained a reference waveform, we can check the accuracy
of the heterodyned likelihood by comparing its value to the original likelihood
at several points in the parameter space. We can choose the number of bins such
that the difference between the values of the two likelihoods is smaller than
chosen tolerance threshold.

@@

Given Eq.~\eqref{eq: loglikelihood} and the prior, one can evaluate the
posterior density, Eq.~\eqref{eq: bayes}, over the entire parameter space of
interest to obtain the most probable set of values that are consistent with the
data. However, directly sampling this posterior quickly becomes intractable as
the dimensionality of the parameter space increases beyond a few dimensions.
Markov chain Monte Carlo \cite{gelmanbda04} is a common method employed to
generate samples from the target posterior when direct sampling is not
possible.

In MCMC, the posterior distribution is approximated by a Markov chain that
eventually converges to the target distribution \cite{10.1214/aos/1176325750}.
The chain is constructed by iteratively proposing a new point in the
parameter space based on the current location of the chain. The proposed point
is accepted with a probability that is usually set to be proportional to the
ratio of the posterior density evaluated at the proposed point and the current
point. The chain can either accept the proposal and move to the new location,
or reject the proposal and stay at the current location. This process is
repeated until the chain converges to the target distribution. The samples
generated by the chain are then used as a fair sample to estimate the
quantities of interest, such as the mean and credible intervals of the source
parameters. In practice, since we do not know the target distribution ahead of
time, the MCMC process is usually repeated until a certain criterion is met,
such as a Gelman-Rubin convergence statistic \cite{10.2307/2246093} lower than
certain threshold, or simply after a fixed number of iterations.

@@

However, as we often deal with high-dimensional problems, even the optimally
tuned Gaussian transition kernel does not guarantee good performance. In order
to have a reasonable acceptance rate, the variance of the Gaussian has to be
smaller in a higher dimensional space, which means that the transition kernel
will generally make smaller and smaller steps as we increase the dimensionality
of the problem \cite{2017arXiv170102434B}.

Transition kernels that leverage gradient information of the target
distribution can help address this issue of shortening steps in a high
dimensional space. Instead of proposing a new point by drawing from a Gaussian,
one can use the gradient evaluated at the current location to propose a new
point, so that the evolution of the chain is preferentially directed to regions
of higher probability. For example, the Metropolis-adjusted Langevin algorithm
(MALA) \cite{10.2307/2346184} places a unit Gaussian at the tip of the gradient
vector at the current position,
\begin{align}
    \mathbf{x} = \mathbf{x_0} + \tau \nabla\log{p(\mathbf{x_0})} + \sqrt{2\tau}N(0,\mathbf{I})\, ,
\end{align}
where $\tau$ is a step size chosen during the tuning stage. Compared to a
Gaussian centered at the current location, the MALA transition kernel is more
likely to propose a point in the higher posterior density region because of the
gradient term, which helps boost the acceptance rate.
Hamiltonian Monte Carlo \cite{2017arXiv170102434B} is another gradient-based
algorithm that has been explored for neutron star inspirals
\cite{Bouffanais:2018hoz}.

While transition kernels that use gradient information can help improve the
acceptance rate, computing the gradient of the posterior density function
introduces an additional computational cost, which is not necessarily
beneficial in terms of sampling time. If one wants to compute the gradient
through finite differencing, the additional computational cost goes as at least
${\sim} \mathcal{O}(2n)$, where $n$ is the dimension of the problem. On the
other hand, schemes like \textsc{Jax}
\cite{jax2018github} allow us to compute the gradient of the likelihood
function with respect to the parameters through automatic differentiation,
which gives the gradient information down to machine precision at around the
same order of time compared to evaluating the posterior itself. Thus, having
access to gradient information through automatic differentiation is crucial to
making gradient-based transition kernels favorable in terms of computing cost.

To leverage automatic differentiation, the entire posterior function must be
implemented within \textsc{Jax} or a similar framework; this includes the
likelihood and, therefore, the waveform approximant.  This means that the
development of differentiable approximants is essential for leveraging
gradient-based sampling in GW applications. We currently make use of the
waveforms implemented in \textsc{ripple} \cite{ripplepaper}.

@@

% Problem with just HMC or MALA
While gradient-based samplers have been shown to outperform gradient-free
algorithms in many practical applications, there remain classes of problems
that most gradient-based samplers do not solve well. For example, first-order
gradient-based algorithms struggle with target distributions that exhibit
locally-varying correlations, since they assume a single mass matrix that does
not depend on the location of the chain by construction
\cite{2017arXiv170102434B}.\footnote{Sampling algorithms that use higher order
derivatives such as manifold-MALA and Riemannian-HMC \cite{RMHMC} can in
principle handle local correlations in the target distribution; however, they
often encounter instabilities when used in real-life applications, so their use
is a rare practice.} Another example is multi-modality: if there are multiple
modes in the target distribution, individual chains will likely be trapped in
one mode and take an extremely long time to transverse between the modes
\cite{2018arXiv180803230M}.  This means that the relative weights between modes
will take much longer to estimate than the shape of each mode.

@@

% Crux of normalizing flow
All the above issues can be mitigated by normalizing flows.  Normalizing flows
are a technique based on neural networks that aims at learning a mapping from a
simple distribution, such as a Gaussian, to a complex distribution, often given
in the form of samples \cite{2019arXiv190809257K, 2019arXiv191202762P}. Once
the network is trained, one can evaluate the probability density of the complex
distribution and sample from it very efficiently, by first evaluating the
simple distribution and then applying the learned mapping. The core equation
of normalizing flows is the coordinate transformation of probability
distributions via a Jacobian, as given by
\begin{align}
    p_x(X) = p_z(Z) \left| \frac{\partial f}{\partial z}\right|^{-1},
\end{align}
where $p_x(X)$ is the complex target distribution, $p_z(Z)$ is the simple
latent distribution and $f$ is an invertible parameterized transform that
connects the two distributions, $x = f(z)$, to be learned by the normalizing
flow. See \cite{2019arXiv190809257K, 2019arXiv191202762P} for a detailed
discussion of the algorithm.

@@

Note that every time we retrain the network, we are breaking the Markov
properties since we are changing the proposal distribution. To produce final
samples that can be used to estimate target quantities, one has to freeze the
normalizing flow model and not retrain during the final sampling phase in order
to satisfy the detailed balance condition. We use the package \textsc{flowMC}
\cite{2022arXiv221106397W,Gabrie:2021tlu}, with MALA as the gradient-based sampler. The
pseudocode of the algorithm is given in Algorithm \ref{alg:cap}.

@@

To demonstrate the robustness of our pipeline, we use it to recover the
parameters of a set of simulated signals injected into different instantiations
of synthetic stationary Gaussian noise. Then we run our pipeline on the
simulated data, and determine the credible interval at which the true
parameters of the injected signals are recovered. From the set of credible
values, we can check whether the truth lies within a certain credible interval
at the expected frequency: if our pipeline is working as expected, we should
find that the true parameters lie within $x\%$ credible interval $x\%$ of the
time, e.g., the true value should lie within the $50\%$ credible interval
$50\%$ of the time. In other words, the recovered percentiles of the true
parameters should be uniformly distributed. Deviation from this behavior would
suggest the pipeline is either over-confident or too conservative
\cite{Cook2006,Talts2018}.

% State the distribution of injected population
We sample 1200 events from the distribution of parameters detailed in Table
\ref{tab:parameters}; the same distributions are used as the prior in the PE
process.  We simulate signals over 16 s of data, with a minimum frequency
cutoff of 30 Hz and a sampling rate of 2048 Hz. We draw noise from a set of
projected design PSDs for the LIGO Hanford, LIGO Livingston
(\texttt{SimNoisePSDaLIGOZeroDetHighPower}) and Virgo
(\texttt{SimNoisePSDAdvVirgo}) detectors
\cite{lalsuite,Shoemaker:T0900288,2012arXiv1202.4031M}. For both injection and
recovery, we make use of the \textsc{IMRPhenomD} waveform \cite{Khan:2015jqa}
via the fully-differentiable implementation presented in the \textsc{ripple}
package \cite{ripplepaper}. We use a neural spline flow model \cite{2019arXiv190604032D}
with 10 layers, each with 128 hidden units, and 8 bins per layer as our
normalizing flow model.

@@

To further quantify how well our result agrees with a uniform distribution, we
can compute the Kolmogorov-Smirnov $p$-values for each marginalized
distribution \cite{}.  A low $p$-value (with a threshold often
chosen to be $p = 0.05$) could indicate that our result is in tension with a uniform
distribution. The $p$-values obtained for each parameter are shown in the
legend of Fig.~\ref{fig:ppplot}. Most of them are well above the $p = 0.05$
threshold, except for $\psi$, which is close to the threshold. Once
again, assuming these $p$-values are drawn from a uniform distribution, given
11 draws (the number of parameters in our inference), it is not abnormal to
have one of the parameters lying slightly outside the threshold. To assess
whether this is expected, we can compute the combined $p$-value for these 11
parameters, and find it to be $p = 0.70$.  This shows our inference pipeline
performs properly on simulated data at a similar level as standard tools
\cite{Veitch:2014wba,Romero-Shaw:2020owr}.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply
it to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors
shown in Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz
starting at 20 Hz for GW150914, and 128 s of data sampled at 4096 Hz starting
at 23 Hz for GW170817; strain data and PSDs for both events are fetched from
GWOSC \cite{GWOSC}. We use the same normalizing flow model as the
injection-recovery study. For our specific choice of sampler settings, we
produce ${\sim}$2500 and 3500 \emph{effective samples}\footnote{Effective
samples here refers to the number of independent samples, which is the total
number of generated samples divided by their correlation length; we compute the
effective sample size using \textsc{arviz}
\cite{arviz_2019},\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}.
} for GW150914 and GW170817 respectively. Running on an \textsc{Nvidia} A100
GPU, the wall time for both events is around 3 minutes. Most of this time is
spent on just-in-time (JIT) compilation of the code; the actual sampling time
is only ${\sim}40$ s. We pre-compute the reference waveform parameters used to
hetetrodyne the likelihood for the two events, which is omitted in the
wall-time calculation.

\begin{figure}
    \includegraphics[width=0.99\linewidth]{figures/GW150914.pdf}
    \caption{
      GW150914 posterior computed by our code (blue) and \textsc{Bilby} (orange).
    }
    \label{fig:GW150914}
    \script{plot_GW150914.py}
\end{figure}

\begin{figure}
\includegraphics[width=0.99\linewidth]{figures/GW170817.pdf}
\caption{
    GW170817 posterior computed by our code (blue) and \textsc{Bilby} (orange).
}
\label{fig:GW170817}
    \script{plot_GW170817.py}
\end{figure}

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\cite{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files included in our data release \cite{release}.  We
carry out these runs using \textsc{parallel Bilby} (\textsc{pBilby})
\cite{Smith:2019ucc} to distribute the computation over 400 Intel Skylake CPUs
for each event.  For the specific settings chosen, the wall-time duration of
each run was ${\sim}2$ h for GW150914 and ${\sim}1$ day for GW170817.

@@

\begin{table}[hbt!]
    \begin{center}
    \variable{output/js_table.tex}
    \caption{JSD values for the marginalized distribution of each parameter for
    GW150914 and GW170817 between our code and \textsc{Bilby}. The bold values
    indicate the parameters with the largest JSD.}
    \label{tab:JSD}
    \end{center}
\end{table}
